Explorations to Optimize the Parareal Algorithm \\to Solve ODEs in Given Applications

### 2014-12-10 paper.tex
added more explanations about figures

Another obstacle to trying to parallelize numerical methods is that the methods
are inherently serial in time---in that evaluation of the next time point $n+1$
depends on the previous values, say those at time $n-1$ and $n$. This setup
does not allow for parallelism. The only way to incorporate parallelism is to
create schemes that have parallelizable components---such as an update or a
refinement built on top of a more basic method first computed in serial. As will be discussed and analyzed in the
methodology section, one such scheme would be the parareal algorithm. As an
overview, the parareal algorithm first does a coarse (low-order and fast)
approximation, then seeks to iteratively correct using smaller, more refined
numerical methods done in parallel from approximated starting points
approximated by the coarse solution.

@@

The parareal algorithm, developed by Lions, Maday and Turinici in
2001~\cite{lions2001parareal}, is a generalized algorithm which allows for
parallelization in time. It does so by using a cheaper ($g_{\Delta t}$)---a
lower order or lesser resolution approximation first, which it then corrects by
evaluating segments with a finer method $g_{\textnormal{fine}}$. The correction
step from the approximation can be done in parallel. Multiple iterations can be
performed to achieve an error provably asymptotically equal to a full serial
computation of $g_{\textnormal{fine}}$. The exact number of iterations required
to achieve certain levels of convergence depends on the problem and must be
tuned top optimize for speedup for a given system. In summary, the algorithm is
$k$ repetitions of a correction process which requires running updated by a
finer method run in parallel for subsections.

The beauty of the parareal method is that it can be adapted into many different versions with different choices of the $g_{\Delta t}$ and $g_{\textnormal{fine}}$ solution operators. The $g_{\textnormal{fine}}$ solution is a more computationally expensive method but assumed to have lesser error.  Most of the time this will be a higher order method, but can also be a greater sampling of the same order method used in the coarse version.  The exact optimum pairing is problem-specific and the code is written so that a different ``n-order method step'' function could be written and then replaced into the code. We experiment with testing different pairs of $g_{\Delta t}$ and $g_{\textnormal{fine}}$ to see how influential the choice of solution scheme is.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.75\linewidth]{parareal_visualization.png}
\caption{Graphical representation of the parareal scheme.}
\label{fig:pararealvis}
\end{center}
\end{figure}

@@

With the assumptions of the previous section, this parareal method will
approach, with large enough $k$ (correction iterations) to approach the error of
the fine method. However, there is a time vs. accuracy tradeoff. Let $Q$ be the
\emph{Quality Factor} for $g_{\Delta t}$ and $g_{\textnormal{fine}}$. Suppose
that, for constant number of processors and $\Delta t$, $g_{\Delta t}$ runs in
time $T$. Then, $g_{\textnormal{fine}}$ runs in time $QT$. With too large of a
$k$ and a lower $Q$, the time for the Parareal could potentially take longer
than the direct serial computation. We note that there could be other
relationships between the time of the coarse and fine schemes other than
multiplication by a constant term.

@@

Let us consider the ODE $du/dt = \lambda u$. Let $g_{\textnormal{fine}}(t_n,
u_n) = \bar{g_{\textnormal{fine}}}u_n$ and $g_{\Delta t}(t_n, u_n) =
\bar{g_{\Delta t}} u_n$. As shown by Staff et al. \cite{staff2005stability}, the parareal method becomes

@@

The first $t$ seconds comes from the first coarse approximation, without which the parareal algorithm degenerates to $g_{\Delta t}$.  Then for each of the $k$ correction iterations, the $t$ term comes from $n$, element by element evaluations of  $g_{\Delta t}$ followed by a parallel $\frac{1}{N}$ evaluation of $g_{\textnormal{fine}}$. In order for there to be a speedup relative to the fine operator, we can rearrange to get:

@@

We first try to express $B$ as function of the quality factor $Q$ and the number of correction steps of the parareal $k$.

@@

It is important to note that this speedup is measured with respect to the parareal agorithm done in parallel. This does not compare the Parareal to doing $g_{\textnormal{fine}}$ but rather shows how given infinite processors, our speedup will approach certain asymptotic bounds. We do see however large constants in the denominator which means that we will get better speedup as we increase n past a certain size.

@@

With embarassingly parallel problem setups, we can optimally achieve up to $n$ times speedup with $n$ processors though this number if quite ideal and will be very difficult to approach (large scaling will have different memory management and greater communication overheads and costs).

@@

We wished to explore certain tests that would demonstrate the effects of changing various parameters be it from the solution schemes in the parareal to the exact system we were trying to solve.

@@

@@

@@

To demonstrate parallelism by space, we solve the two-dimensional wave equation by partioning a two-dimensional space domain amongst multiple processors. Specifically, we look to numerically solve

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\linewidth]{decomp.png}
\caption{Parallelization by space schematic.}
\label{fig:spaceparallel}
\end{center}
\end{figure}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|} \hline
($N_x, N_y$) & $P_x$ & $P_y$ & Time (sec) & Speedup \\ \hline
%(128, 128) & 2 & 2 & (still queued on Odyssey) & \\ \hline
%(128, 128) & 4 & 2 & (still queued on Odyssey) & \\ \hline %check
%(128, 128) & 4 & 4 & (still queued on Odyssey) & \\ \hline
%(128, 128) & 8 & 8 & (still queued on Odyssey) & \\ \hline
(256, 256) & 2 & 2 & 2.57 & 3.07\\ \hline %real
(256, 256) & 4 & 2 & 11.32 & \\ \hline %check
(256, 256) & 4 & 4 & 13.65 & \\ \hline %check
(256, 256) & 8 & 8 & (still queued on Odyssey)& \\ \hline %check17.44
(512, 512) & 2 & 2 & 39.65 & 1.61\\ \hline %real
(512, 512) & 4 & 2 & 14.46 & 4.42\\ \hline
(512, 512) & 4 & 4 & (still queued on Odyssey)&\\ \hline %check
(512, 512) & 8 & 8 & (still queued on Odyssey)&\\ \hline %check
\end{tabular}
\end{center}

@@

We see that (HOPE IT WORKS WELL OR ELSE...) Most of the further work would be in
optimization. We are trying to measure performance in Python, which is not the
ideal benchmarking language for efficiency and speedups. A port over to C++
using the MPI libraries in C++ rather than the mpi4py libraries in python would
be ideal. Other optimizations mentioned above could also be implemented. To
further explore the parareal algorithm would involve many test cases and to see
how the parareal algorithm compares to other approaches---since the actual time
and iterations necessary depend on the system of ODEs to solve.

@@

On the other hand, parallelism by space already requires a more problem-specific
design. Parallelism by space is intuitive, and conceptually, is embarrassingly parallel and easier to create. However, the
limitations are in the requirement for the code to be designed for the
problem---how to divide by space so as to allow for the maximum number of even
divisions. The lack of generality is not as beautiful as the parareal algorithm
but may lend itself to faster speedups.

@@

All in all, we have implemented and begun to explore some techniques of looking
for strong and weak scaling efficiencies for solving ODEs. The parareal
algorithm is beautiful in its theoretical advantages---in terms of stability,
error and efficiency. However, the method, as a very generalized method,
requires deeper analysis for the specific problem. The main variable will be in
the difference in the coarse and fine methods (be it lower and higher order or a
higher and lower resolution for the same method). The tradeoffs that must be
computed and optimized for would be problem specific.

@@

Bibtex Citations for the Slides, and for \url{https://www.sharcnet.ca/help/index.php/Measuring_Parallel_Scaling_Performance}

>>>>

Another obstacle to trying to parallelize numerical methods is that the methods
are inherently serial in time---in that evaluation of the next time point $n+1$
depends on the previous values, say those at time $n-1$ and $n$. This setup
does not allow for parallelism. The only way to incorporate parallelism is to
create schemes that have parallelizable components---such as an update or a
refinement built on top of a more basic method first computed in serial. As will be discussed and analyzed in the
methodology section, one such scheme would be the Parareal algorithm. As an
overview, the Parareal algorithm first does a coarse (low-order and fast)
approximation, then seeks to iteratively correct using smaller, more refined
numerical methods done in parallel from approximated starting points
approximated by the coarse solution.

@@

The Parareal algorithm, developed by Lions, Maday and Turinici in
2001~\cite{lions2001parareal}, is a generalized algorithm which allows for
parallelization in time. It does so by using a cheaper ($g_{\Delta t}$)---a
lower order or lesser resolution approximation first, which it then corrects by
evaluating segments with a finer method $g_{\textnormal{fine}}$. The correction
step from the approximation can be done in parallel. Multiple iterations can be
performed to achieve an error provably asymptotically equal to a full serial
computation of $g_{\textnormal{fine}}$. The exact number of iterations required
to achieve certain levels of convergence depends on the problem and must be
tuned top optimize for speedup for a given system. In summary, the algorithm is
$k$ repetitions of a correction process which requires running updated by a
finer method run in parallel for subsections.

The beauty of the Parareal method is that it can be adapted into many different versions with different choices of the $g_{\Delta t}$ and $g_{\textnormal{fine}}$ solution operators. The $g_{\textnormal{fine}}$ solution is a more computationally expensive method but assumed to have lesser error. Most of the time this will be a higher order method, but can also be a greater sampling of the same order method used in the coarse version. The exact optimum pairing is problem-specific and the code is written so that a different ``n-order method step'' function could be written and then replaced into the code. We experiment with testing different pairs of $g_{\Delta t}$ and $g_{\textnormal{fine}}$ to see how influential the choice of solution scheme is.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.75\linewidth]{parareal_visualization.png}
\caption{Graphical representation of the Parareal scheme.}
\label{fig:pararealvis}
\end{center}
\end{figure}

@@

With the assumptions of the previous section, this Parareal method will
approach, with large enough $k$ (correction iterations) to approach the error of
the fine method. However, there is a time vs. accuracy tradeoff. Let $Q$ be the
\emph{Quality Factor} for $g_{\Delta t}$ and $g_{\textnormal{fine}}$. Suppose
that, for constant number of processors and $\Delta t$, $g_{\Delta t}$ runs in
time $T$. Then, $g_{\textnormal{fine}}$ runs in time $QT$. With too large of a
$k$ and a lower $Q$, the time for the Parareal could potentially take longer
than the direct serial computation. We note that there could be other
relationships between the time of the coarse and fine schemes other than
multiplication by a constant term.

@@

Let us consider the ODE $du/dt = \lambda u$. Let $g_{\textnormal{fine}}(t_n,
u_n) = \bar{g_{\textnormal{fine}}}u_n$ and $g_{\Delta t}(t_n, u_n) =
\bar{g_{\Delta t}} u_n$. As shown by Staff et al. \cite{staff2005stability}, the Parareal method becomes

@@

The first $t$ seconds comes from the first coarse approximation, without which the Parareal algorithm degenerates to $g_{\Delta t}$. Then for each of the $k$ correction iterations, the $t$ term comes from $n$, element by element evaluations of $g_{\Delta t}$ followed by a parallel $\frac{1}{N}$ evaluation of $g_{\textnormal{fine}}$. In order for there to be a speedup relative to the fine operator, we can rearrange to get:

@@

We first try to express $B$ as function of the quality factor $Q$ and the number of correction steps of the Parareal $k$.

@@

It is important to note that this speedup is measured with respect to the Parareal algorithm done in parallel. This does not compare the Parareal to doing $g_{\textnormal{fine}}$ but rather shows how given infinite processors, our speedup will approach certain asymptotic bounds. We do see however large constants in the denominator which means that we will get better speedup as we increase n past a certain size.

@@

With embarrassingly parallel problem setups, we can optimally achieve up to $n$ times speedup with $n$ processors though this number if quite ideal and will be very difficult to approach (large scaling will have different memory management and greater communication overheads and costs).

@@

We wished to explore certain tests that would demonstrate the effects of changing various parameters be it from the solution schemes in the Parareal to the exact system we were trying to solve.

@@

As seen in Figures~\ref{fig:err_v_k} and~\ref{fig:err_v_q}, the error remains
constant for the parallel versions of the algorithm even as we increase the
number of correction steps and the quality factor. This is not exactly what we
expect. We predicted that the error would decrease as we did more correction
steps, or if we increased the quality of the $g_{\textnormal{fine}}$ operator.

@@

Figure~\ref{fig:run_v_k} shows the running time (wall clock) as a function of
the number of correction steps. For the parallel implementations, we see an
increase in running time as we increase the number of correction steps. This is
to be expected, since increasing the number of correction steps increases the
number of iterations of our algorithm. Therefore, even though we proved before
that as the number of iterations $k \to N$, where $N$ is the number of time
points, the order of convergence of the algorithm converges to the order of
accuracy of $g_{\textnormal{fine}}$, this comes at the expense of increasing
computational cost.

\subsection{Comparison to Serial with $g_{\textnormal{fine}}$ as Higher Order
Methods}

We considered using a higher order $g_{\textnormal{fine}}$ in our
implementation, but we were unable to show higher order convergence of the error
when we tested our method serially.

@@

To demonstrate parallelism by space, we solve the two-dimensional wave equation by partitioning a two-dimensional space domain amongst multiple processors. Specifically, we look to numerically solve

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\linewidth]{decomp.png}
\caption{Parallelization by space schematic.}
\label{fig:spaceparallel}
\end{center}
\end{figure}

@@

\begin{center}
\begin{tabular}{|c|c|c|c|c|} \hline
($N_x, N_y$) & $P_x$ & $P_y$ & Time (sec) & Speedup \\ \hline
%(128, 128) & 2 & 2 & (still queued on Odyssey) & \\ \hline
%(128, 128) & 4 & 2 & (still queued on Odyssey) & \\ \hline %check
%(128, 128) & 4 & 4 & (still queued on Odyssey) & \\ \hline
%(128, 128) & 8 & 8 & (still queued on Odyssey) & \\ \hline
(256, 256) & 2 & 2 & 2.57 & 3.07\\ \hline %real
(256, 256) & 4 & 2 & 11.32 & \\ \hline %check
(256, 256) & 4 & 4 & 13.65 & \\ \hline %check
(256, 256) & 8 & 8 & 17.44 & \\ \hline %check
(512, 512) & 2 & 2 & 39.65 & 1.61\\ \hline %real
(512, 512) & 4 & 2 & (still queued on Odyssey)& \\ \hline
(512, 512) & 4 & 4 & (still queued on Odyssey)&\\ \hline %check
(512, 512) & 8 & 8 & (still queued on Odyssey)&\\ \hline %check
\end{tabular}
\end{center}

@@

We see that (HOPE IT WORKS WELL OR ELSE...) Most of the further work would be in
optimization. We are trying to measure performance in Python, which is not the
ideal benchmarking language for efficiency and speedups. A port over to C++
using the MPI libraries in C++ rather than the mpi4py libraries in python would
be ideal. Other optimizations mentioned above could also be implemented. To
further explore the Parareal algorithm would involve many test cases and to see
how the Parareal algorithm compares to other approaches---since the actual time
and iterations necessary depend on the system of ODEs to solve.

@@

On the other hand, parallelism by space already requires a more problem-specific
design. Parallelism by space is intuitive, and conceptually, is embarrassingly parallel and easier to create. However, the
limitations are in the requirement for the code to be designed for the
problem---how to divide by space so as to allow for the maximum number of even
divisions. The lack of generality is not as beautiful as the Parareal algorithm
but may lend itself to faster speedups.

@@

All in all, we have implemented and begun to explore some techniques of looking
for strong and weak scaling efficiencies for solving ODEs. The Parareal
algorithm is beautiful in its theoretical advantages---in terms of stability,
error and efficiency. However, the method, as a very generalized method,
requires deeper analysis for the specific problem. The main variable will be in
the difference in the coarse and fine methods (be it lower and higher order or a
higher and lower resolution for the same method). The tradeoffs that must be
computed and optimized for would be problem specific.

@@

\section{Appendix}
\subsection{Explanation of Code Files}
